Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-10-29 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response. Xylem architecture plays a role, with diffuse porous species tending to be more drought sensitive than ring-porous species [@kannenberg_linking_2019] *(Elliot et al. 2015-doi: 10.1111/gcb.13045, Friedrichs et al. 2009-doi:10.1007/s00468-009-0315-2)*, but does not differentiate species beyond broad classes. To link species' level drought responses to fundamental physiological characteristics, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009]*(Hoffmann et al. 2011- doi: 10.1111/j.1365-2486.2011.02401.x)* and other forest biomes around the world (*Greenwood et al. 2007- doi: 10.1111/ele.12748*). However, the direction of response is not always consistent; wood density correlated negatively with drought tree drought performance in a broadleaf deciduous forest in the southeast US *(Hoffmann et al. 2011)* but positively across a global scale (*Greenwood et al. 2007- doi: 10.1111/ele.12748*). Leaf hydraulic traits with more direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019]. The ability of these hydraulic traits to tree performance under drought remains untested.

A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts. No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018]. However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts. (**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment. The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; Stovall; **MORE REFS**). Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited. While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**). There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019]. [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
(**PELTIER**)

Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses. 
First, we focus on the role of tree size and its interaction with microenvironement. 
We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow observed patterns in relation to wood density and specific leaf area, but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2018 (**confirm**), are available through the ForestGEO data portal(www.forestgeo.si.edu).

\noindent\hrulefill 
# ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

@@

We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity (ANPP), which together comprised 97% of study plot ANPP between 2008 and 2013 [@helcoski_growing_2019]. Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  *Ryan submitted the data to ITRDB but I don't think its posted yet. We should also cite GitHub/Zenodo here. I'll come back to that.*

@@

$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$

Here, *Y* denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter without bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH. Additional species were used here than used in the final analysis to establish base averages. (**do we want to keep this or take out juni/fram from table?**)

Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018]. Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013]. Species-specific height allometries were developed (Table S2). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.

>>>>

A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response. Xylem architecture plays a role, with diffuse porous species tending to be more drought sensitive than ring-porous species [@kannenberg_linking_2019] *(Elliot et al. 2015-doi: 10.1111/gcb.13045, Friedrichs et al. 2009-doi:10.1007/s00468-009-0315-2)*, but does not differentiate species beyond broad classes. Commonly measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009]*(Hoffmann et al. 2011- doi: 10.1111/j.1365-2486.2011.02401.x)* and other forest biomes around the world (*Greenwood et al. 2007- doi: 10.1111/ele.12748*). However, the direction of response is not always consistent; wood density correlated negatively with drought tree drought performance in a broadleaf deciduous forest in the southeastern United States *(Hoffmann et al. 2011)* but positively or with no effect at a global scale (*Greenwood et al. 2007- doi: 10.1111/ele.12748; Anderegg et al. 2016- doi: 10.1073/pnas.1525678113*). Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance (*Anderegg et al. 2016*) but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019]. The ability of these hydraulic traits to tree performance under drought remains untested.

A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts. No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018]. However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment. The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019] *Anderegg et al. 2016*). Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited. While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**). There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019]. [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.

Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses. 
First, we focus on the role of tree size and its interaction with microenvironement. 
We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; *Hoffmann et al. 2011*) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2018 (**Ian, confirm**), are available through the ForestGEO data portal (www.forestgeo.si.edu).

\noindent\hrulefill 
 ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

@@

We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019]. Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  *Ryan submitted the data to ITRDB but I don't think its posted yet. We should also cite GitHub/Zenodo here. I'll come back to that.*

@@

$$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$$

Here, $Y$ denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter without bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.

Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Species-specific height allometries were developed (Table S2). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.
